Sketch Recognition using Particle Swarm Optimizations

### 2009-06-06 paper2/Paper2.tex

\author[ERI]{Maha El Meseery\footnote{Crossponding author}}
\author[ERI]{Mahmoud Fakhr El Din}
\author[ERI]{Samia Mashali}
\author[FAC]{Magda Fayek}
\author[FAC]{Nevin Darwish}
\address[ERI]{Signals Processing Group, Computers and Systems Department, 
  Electronic Research Institute, Cairo, Egypt\\melmeseery@eri.sci.eg, mafakhr@mcit.gov.eg, samia@eri.sci.eg}
\address[FAC]{Computer Engineering Department,
	Faculty of Engineering,Cairo University, Cairo, Egypt\\
	magdafayek@gmail.com,ndarwish@ieee.org}
\begin{abstract}
%%This study 
Sketches and drawings are widely used as simple methods of expressing thought and designs. The goal of this paper is developing a sketch recognition system that facilitates using sketches in computer systems. The system uses Particle Swarm Optimization (PSO) algorithm to optimally segment the strokes the user draws into meaningful geometric primitives.  These geometric primitives are grouped to form symbols which are further identified using a Support Vector Machines (SVM) classifier. Two versions of PSO segmentation algorithms were implemented. Experiments were conducted on three different dataset, simple presentation (Hs-DS), Electrical Design (EL-DS) and Logic Design (LD-DS) symbols. Results show that both PSO algorithms produce better segmentation than most current systems. This segmentation improvement contributed in getting more accurate recognition results on all the datasets tested. The results prove that the system is effective specially in datasets with low diversity in shapes, i.e., shapes are very simillar, as in Logic Design dataset (LD-DS). %The results prove the effectiveness of the proposed system on various dataset, especially datasets with low shape diversity as in Logic Design dataset (LD-DS).% effectiveness of the proposed system on various dataset, especially datasets with low shape diversity as in Logic Design dataset (LD-DS). %Result show that the system is especially effictive on dataset with % sketch are divided  Users draw symbols and sketchs using a set of pen strokes. 
%%Sketch recognition is defined as the process of identifying symbols that users draw using single or multiple strokes. Usually, users draw strokes using pens. The sketch recognition system immediately interprets their strokes as objects that can be easily manipulated. This paper uses Particle Swarm Optimization (PSO) to divide the strokes the user draws into meaningful geometric primitives. These geometric primitives are grouped to formulate symbols which are further identified. The results show that using PSO improves segmentation results which guide the symbol recognition phase. As for recognition we use Support Vector Machines (SVM) classifier which further improves the final recognition accuracy.

@@

\begin{keyword}
%% keywords here, in the form: keyword \sep keyword
Symbol Recognition, Particle Swarm Optimization, Sketch Recognition, Geometerical Features, Polygon approximation
%% PACS codes here, in the form: \PACS code \sep code

@@

%\begin{abstract}
%%This study 
%Sketches and drawings are widely used as a simple method of expressing thought and designs. The goal of this paper is developing a sketch recognition system that facilitates using sketches in computer systems. The system uses Particle Swarm Optimization (PSO) algorithm to optimally segment the strokes the user draws into meaningful geometric primitives.  These geometric primitives are grouped to formulate symbols which are further identified using a Support Vector Machines (SVM) classifier. Two different segmentation algorithms were evaluated and results show that both algorithms improve the final recognition results. Experiments were conducted on three different dataset, simple presentation (Hs-DS), Electrical Design (EL-DS) and Logic Design (LD-DS) symbols. Results show that the effectiveness of the proposed system on various dataset especially on datasets with low shape diversity as in Logic Design dataset (LD-DS).  % sketch are divided  Users draw symbols and sketchs using a set of pen strokes. 
%%Sketch recognition is defined as the process of identifying symbols that users draw using single or multiple %strokes. Usually, users draw strokes using pens. The sketch recognition system immediately interprets their %strokes as objects that can be easily manipulated. This paper uses Particle Swarm Optimization (PSO) to divide %the strokes the user draws into meaningful geometric primitives. These geometric primitives are grouped to %formulate symbols which are further identified. The results show that using PSO improves segmentation results %which guide the symbol recognition phase. As for recognition we use Support Vector Machines (SVM) classifier %which further improves the final recognition accuracy.  
%\end{abstract}
\pagenumbering{arabic}
\section{Introduction}

@@

Sketch recognition can be applied either using  Pattern Recognition (PR) based method or Artifical Intelligance (AI) based method. In AI based methods, Symbols are represented using hierarchical levels of simpler geometric shapes or primitives and spatial relations between them. Systems that employ such representation must first pre-process. strokes to break them into geometric primitives. Recognition can then is treated as an AI problem which can be solved using methods like sub graph matching, bayes belief network, and constrain statisfaction. A Hierarchical shape descriptions was widely used in sketch systems \cite{HierarchicalParsing7,SketchRead2007}.  %Ladder30,AlvaradoFreedom42

Recognition based on AI methods as applied in \cite{SketchRead2007} has the principle problem of computational complexity of the matching between shape description and geometric primitives produced in the segmentation process. Natural sketch are difficult to be segmented and grouped to create reliable fragment of geometric primitives. This is partially due to the increase of noise and other phenomena such as overlapped, over traced strokes and sharing of strokes between different symbols. Since these systems are dependable on the segmentation results, they either perform recognition using undependable segmentation results or consider different ways to segment the input sketch. This is because they assume that each stroke plays a role in the structure of the shape. 
%The role of which each stroke plays in the structure of the sketched symbol is used to determine shapes in stroke based methods.  
%Gestures are pen strokes drawn by the user that can immediately recognized by the system.
%role of which each stroke plays in the structure of the sketched symbol is used to determine shapes 
%. discribing the role  how user draws individal storkes  properties that discribe how users draw individual strokes in drawing the symbol.  to . The identification of individual stroke is used to determine  The symbols are recognized by identifing the individual strokes.
In PR based methods, systems compute properties that discribe symbols then use a classification method to recognize symbols. The properties used can either be based on gesture properties, global shape properties or image based properties. Gesture properties are set of features that represent how the user draws a symbol not how it looks like (e.g., identify a regtangle when a user draws a L corner). In gesture properties approcah, systems identify symbols by using gesutre properties of some of its  underlying strokes. System that uses gestures properties assume that users draw symbols in specific order and style \cite{gestureexample12,aideddesgin22}. In global shape properties approach, systems investigate general shape properties and its underlying strokes. Using shape properties (e.g., ratio of the area of convex hull to the area of shape) relaxes the assumption and the role of each stroke in the symbol but it does not truly represent the appearance or fine details of the symbol \cite{DiagramOfflineConvexHull,Cali63}. Unlike shape properties based methods, image based methods disregard individual strokes and focus on the overall appearance of shapes. These system convertes the shapes into image and extracte image based features (e.g. local gradient feature)\cite{Oltmans07,imagetrainable48}. In \cite{imagetrainable48}, they used four different distance measures to to match user symbol with the prototype database. The prototype of a shape is represented as an image of the symbol that is normalized and scaled.  However, the problem is that the shape prototype must represent all transformation and variation the input shape can have.   %Image based methods are used in \cite{Oltmans07,imagetrainable48} where they computed a code book of 1000 features to recognize symbols.

@@

In this paper, we choose to implement a hybrid method between shape property, guesture and hierarchical (AI method) based approaches. Our method attempts to avoid many of the problems faced by approaches based on individual guestures or global shape properties while trying to gain the additional information that the stroke based system uses. Firstly, the system uses particle swarm optimization (PSO) to optimally segment user's strokes. Users can draw symbols using any number of strokes in any order they like. The system segments each stroke using two different PSO algorithms. The first segmentation algorithm is based on dividing the digital curve into polygons\cite{PolygonApproximationPSO}. Our algorithm enhances its performance by adding curvature and other local information while segmenting strokes to achieve better results. To handle complex strokes, our system uses a second segmentation PSO algorithm to segment strokes into a set of lines and circular arcs. After segmentation, a feature vector composed of different statistical, geometrical and composite features is deduced. The final classification step uses a SVM classifier to classify symbols into one of the previously trained classes based on the features vector computed.

 %In general, the  appearance based methods use image processing techniques to recognize sketches that the users draw. 
%Sketch recognition can be based on: strokes, global shape properties, and appearance \cite{Oltmans07}. The role of which each stroke plays in the structure of the sketched symbol is used to determine shapes in stroke based methods. Most stroke based systems either represent symbols as gestures or as a  hierarchical shape. Gestures are pen strokes drawn by the user that can immediately recognized by the system. A Hierarchical shape descriptions was widely used in sketch systems \cite{HierarchicalParsing7,Ladder30,AlvaradoFreedom42,SketchRead2007}. Symbols are represented using hierarchical levels of simpler geometric shapes or primitives and spatial relations between them. Systems that employ such representation must first pre-process strokes to break them into geometric primitives. Recognition can hen be treated as sub graph matching problem.

@@

\subsection{Preprocessing}
\label{Prepross}% Agar \cite{polygonfeedback31} mentions that the pointing device (for example the mouse) sampling rate is the reason for this phenomena.
 Speed and time difference information were widely used in sketch recognition systems \cite{earlyprocess}. Figure \ref{fig:orignalStroke} shows an example of an input stroke drawn by users.  Figure \ref{fig:speed2Distance} shows the graph of time difference and the speed for the stroke in Figure \ref{fig:orignalStroke}. It was noted that time difference provides more distinct maxima than the speed information. Agar et al. \cite{polygonfeedback31} mention that the pointing device (e.g., mouse) sampling rate is the reason for this phenomena. The points are sampled at regular time intervals as long as the pointing device is moving. Therefore, there are no new detected points when the pointing device is stationary. This leads to a nearly constant time difference between samples while the pen is moving and large difference while the pen is stationary (Figure \ref{fig:timediff}). Contrary to time difference information, the speed information, where the users draw with variant speeds, the speed information has a lot of noise in the data (Figure \ref{fig:speed2Distance}).  %Similarly, direction information is used as it provides better distinctive maxima for the corners than the curvature information \cite{meanshift10}.% Figure \ref{fig:curvatures} shows the direction and curvature graphs for the stroke drawn in Figure \ref{fig:orignalStroke}

@@

 \begin{figure}
	\centering
			\subfigure[Speed versus Displacement Graph]{	\includegraphics[scale=0.4]{images/vel3.jpg}}
			\hfill
			\subfigure[ Time Difference  versus Displacement Graph] {\label{fig:timediff}\includegraphics[scale=0.4]{images/td3.jpg}}
	\caption{Speed and Time Difference Graphs}  a) speed graph has lot of noise in the data with respect to b) time difference.   The black points denote the location of the possible dominant points $P_{pd}$. Black rectangles represent regions $R_i$ that in a) are lower than the thershold. b) are above the thershold. 
	\label{fig:speed2Distance}
\end{figure}
%is calculated as the angle between two vectors and  where $Q_i=\overrightarrow {P_{i - 1} P_i }$ is the vector from the point $P_{i - 1}$ to point $P_i$ 
Dominant points are characterized by low speed values, high curvature and direction values. However, the dominant points from the curvature information may not be the same points as the dominant points extracted from the speed information. Hence, we compute time difference, direction, speed and curvature of each point along a stroke to ensure that no dominant point is overlooked in this phase. The speed is calculated as $v=\Delta s/\Delta t$ where $t$ is the time difference between two points and $s$ is the length between them. The direction $d$ calculated as the angle between vectors $\overrightarrow {P_{i - 1} P_i }$ and the $x-axis$ and the curvature is considered as the change in direction $d$ with respect to length $s$ i.e. $c= \Delta d/\Delta s$.

@@

\subsection{Segmentation}
\label{seg}
In the segmentation stage, the input stroke is divided into a set of primitives. As shown in Figure \ref{fig:segblock} first an attempt is made to fit the stroke points into a curve or an ellipse using a minimum square error fitting algorithm \cite{ellipsefit}. The ellipse fitting step mainly helps to prevent the system from over segmenting the stroke into multiple lines or curves if the input stroke is an ellipse. If the stroke proved to be an ellipse arc then the segmentation process ends and the system proceeds to the next step. Otherwise, the stroke is passed to two further segmentation algorithms that divide the stroke to either lines or lines and curves. Hence we generate two different segmentations, the system chooses the segmentation with the minimum error. The following sections describe in details the ellipse detection algorithm and the two segmentation algorithms used to divide the stroke. % The input to these algorithms are the stroke points along with the possible dominant points $P_{pd}$ computed during preprocessing. The output is a set of dominant points which are connected by either lines or curves. 
%After two different segmentations are generated, 
%///AllBlockDiagram.jpg  you can change the block to the main block. 
 \begin{figure}
	\centering
		\includegraphics[scale=0.48]{images/blockSmall.jpg}
	\caption{Segmentation Block} 
	\label{fig:segblock}
\end{figure}
\subsection{Ellipse fitting}
%The process starts by computing the center of the stroke bounding box. The bounding box center point is used as the first estimation of the ellipse center. The axes of the ellipse are estimated as $width/2$ and $height/2$ of the stroke bounding box. The least square fitting algorithm is used to minimize the fitting error of the ellipse equation \cite{chernov}.  
This process attempts to fit the stroke points into an ellipse arc; it starts with computing the center of the stroke bounding box. The bounding box center point is used as the first estimation of the center of the ellipse. The axes of the ellipse are estimated as the $width/2$ and $height/2$ of the stroke bounding box where $width$ and $height$ are the width and height of the bonding box. The least square fitting algorithm \cite{chernov} is used to minimize the fitting error of the ellipse Equation (\ref{eq:circleFit})  
\begin{equation}
E = \sum\limits_{i = 0}^N {\frac{{(x_i - x_0 )}}{{a^2 }}^2  + \frac{{(y_i - y_0 )}}{{b^2 }}^2  - 1} 
\label{eq:circleFit}
\end{equation}
where $N$ is number of points in the stroke, $a$ \& $b$ are the length of ellipse axes, $x_0$ \& $y_0$ are the coordinates of the center point and $x_i$ \& $y_i$ are the coordinates of point $i$ in the stroke. A list of new values for $x_0$ , $y_0$ ,$a$ and $b$ are generated randomly from the older values with small increments after each loop.  After few iterations, the final fit error of the estimated ellipse is reported. Another measure is used to compute the efficiency of the final estimated ellipse. Equation \ref{eq:circleError} ensures that the drawn percentage of ellipse is considered. This eliminates fitting a  line into very large ellipse but leaves small ellipses to be fitted as a partial or a full ellipse.

@@

After each iteration of the swarm algorithm (\textsl{AlgS1} and \textsl{AlgS2}), each particle is refined using the following procedures: 
\begin{enumerate}
	\item For each particle $P_i$ each dominant point $P_{ij}$ is checked to find if it was labeled before as a \textit{possible dominant point} $P_{pd}$ (computed as in section \ref{Prepross}). If it was not labeled, the point $P_{ij}$ is moved to the nearest labeled point. This ensures that all of the points generated by the DPSO are possible dominant points $P_{pd}$. 
	\item The particles are tested to make sure that the distance between every two successive dominate point is larger than $min_D$, where $min_D$ is 5\% of the total length of the stroke.  Otherwise, two temperary particles are evaluated by removing either point from the current particle. After that, one of these points is removed based on the error cased by its removal.
	\item Each two successive segments are tested to determine if they can be merged into a single segment. If merging the two segments results in segmentation that has an error value less than or equal to the error before merging, the two segments are merged. This step ensures that the final segmentation of the system has minimum number of segments. 
	% the same or smaller the error
\end{enumerate}
%1) 2)  % If two points are nearer than $min_D$ then one of the points is removed. 
\subsection{Recognition}
\label{sec:Recognition}
After the user draws all strokes of the symbol, the set of un-recognized strokes is grouped together along with their segmentation as input to the feature extraction process. A composite set of features is used to generate a single feature vector. The following list describes each feature in details.
%\begin{description}
\subsubsection{Structural and geometrical Features (FS1)}
  They are features defines the structure of the geometrical symbol such as;   
%\begin{description}
%	 \item[No. Of segments] Number of segments in the symbol. 
%		\item[No. Of primitives] Number of primitives in the symbol. The feature helps when identifying             symbols with mixed geometric primitives like cylinders and callouts.  
%		\item [No. of curves] Number of curves or ellipses in the symbol. 
%		\item [No. of lines] Number of lines in the symbol. 
%		\item [No. of perpendicular lines] Number of perpendicular lines. 
%		\item [No. of parallel lines] Number of parallel lines. 
%		\item [No. of intersections] Number of intersection between lines and curves. 
%		\item [Size Ratio] Ratio between width to height of the symbol.
%		%\item[No. of holes] Number of holes in the symbols. 
%
%\end{description}
 \begin{itemize}
	 \item \emph{Segments:} The number of segments in the symbol.
	 \item \emph{Strokes:} The number of strokes or partial strokes that created the symbols.  
		\item  \emph{Primitives:} The umber of primitives in the symbol. This feature helps when identifying  symbols with mixed geometric primitives such as cylinders and callouts.  
		\item \emph{Curves:} The number of curves or ellipses in the symbol. 
		\item \emph{Lines:} The number of lines in the symbol. 
		\item \emph{Perpendicular}\emph{lines:} The number of perpendicular lines. 
		\item \emph{Parallel}\emph{lines:} The number of parallel lines. 
		\item \emph{Intersections:} The number of intersection between lines and curves. 
		\item \emph{T intersections:} The number of T intersections. 
		\item \emph{L intersections:} The number of L intersections. 
		\item \emph{X intersections:} The number of X intersections.	
		%\item[No. of holes] Number of holes in the symbols. 
\end{itemize}
\subsubsection{Rubine Features (FS2)}
  These features introduced by Rubine \cite{gestureexample12} for single stroke gestures. Such as length of stroke, cosine of starting angle, smoothness of symbol and the properties of the  bounding box of the symbol.

>>>>

\author[ERI]{Maha El Meseery\footnote{Corresponding author}}
\author[ERI]{Mahmoud Fakhr El Din}
\author[ERI]{Samia Mashali}
\author[FAC]{Magda Fayek}
\author[FAC]{Nevin Darwish}
\address[ERI]{Signals Processing Group, Computers and Systems Department, 
  Electronic Research Institute, Cairo, Egypt\\melmeseery@eri.sci.eg, mafakhr@mcit.gov.eg, samia@eri.sci.eg}
\address[FAC]{Computer Engineering Department,
	Faculty of Engineering, Cairo University, Cairo, Egypt\\
	magdafayek@gmail.com, ndarwish@ieee.org}
\begin{abstract}
%%This study 
Sketches and drawings are widely used as simple methods of expressing thought and designs. The goal of this paper is developing a sketch recognition system that facilitates using sketches in computer systems. The system uses Particle Swarm Optimization (PSO) algorithm to optimally segment the strokes the user draws into meaningful geometric primitives.  These geometric primitives are grouped to form symbols which are further identified using a Support Vector Machines (SVM) classifier. Two versions of PSO segmentation algorithms were implemented. Experiments were conducted on three different dataset, simple presentation (Hs-DS), Electrical Design (EL-DS) and Logic Design (LD-DS) symbols. Results show that both PSO algorithms produce better segmentation than most current systems. This segmentation improvement contributed in getting more accurate recognition results on all the datasets tested. The results prove that the system is effective specially in datasets with low diversity in shapes, i.e., shapes are very similar, as in Logic Design dataset (LD-DS). %The results prove the effectiveness of the proposed system on various dataset, especially datasets with low shape diversity as in Logic Design dataset (LD-DS).% effectiveness of the proposed system on various dataset, especially datasets with low shape diversity as in Logic Design dataset (LD-DS). %Result show that the system is especially effective on dataset with % sketch are divided Users draw symbols and sketches using a set of pen strokes. 
%%Sketch recognition is defined as the process of identifying symbols that users draw using single or multiple strokes. Usually, users draw strokes using pens. The sketch recognition system immediately interprets their strokes as objects that can be easily manipulated. This paper uses Particle Swarm Optimization (PSO) to divide the strokes the user draws into meaningful geometric primitives. These geometric primitives are grouped to formulate symbols which are further identified. The results show that using PSO improves segmentation results which guide the symbol recognition phase. As for recognition we use Support Vector Machines (SVM) classifier which further improves the final recognition accuracy.

@@

\begin{keyword}
%% keywords here, in the form: keyword \sep keyword
Symbol Recognition, Particle Swarm Optimization, Sketch Recognition, Geometrical Features, Polygon approximation
%% PACS codes here, in the form: \PACS code \sep code

@@

%\begin{abstract}
%%This study 
%Sketches and drawings are widely used as a simple method of expressing thought and designs. The goal of this paper is developing a sketch recognition system that facilitates using sketches in computer systems. The system uses Particle Swarm Optimization (PSO) algorithm to optimally segment the strokes the user draws into meaningful geometric primitives.  These geometric primitives are grouped to formulate symbols which are further identified using a Support Vector Machines (SVM) classifier. Two different segmentation algorithms were evaluated and results show that both algorithms improve the final recognition results. Experiments were conducted on three different dataset, simple presentation (Hs-DS), Electrical Design (EL-DS) and Logic Design (LD-DS) symbols. Results show that the effectiveness of the proposed system on various dataset especially on datasets with low shape diversity as in Logic Design dataset (LD-DS).  % sketch are divided Users draw symbols and sketches using a set of pen strokes. 
%%Sketch recognition is defined as the process of identifying symbols that users draw using single or multiple %strokes. Usually, users draw strokes using pens. The sketch recognition system immediately interprets their %strokes as objects that can be easily manipulated. This paper uses Particle Swarm Optimization (PSO) to divide %the strokes the user draws into meaningful geometric primitives. These geometric primitives are grouped to %formulate symbols which are further identified. The results show that using PSO improves segmentation results %which guide the symbol recognition phase. As for recognition we use Support Vector Machines (SVM) classifier %which further improves the final recognition accuracy.  
%\end{abstract}
\pagenumbering{arabic}
\section{Introduction}

@@

Sketch recognition can be applied either using Pattern Recognition (PR) based method or Artificial Intelligence (AI) based method. In AI based methods, Symbols are represented using hierarchical levels of simpler geometric shapes or primitives and spatial relations between them. Systems that employ such representation must first pre-process strokes to break them into geometric primitives. Recognition can then is treated as an AI problem which can be solved using methods like sub graph matching, Bayes belief network, and constrain satisfaction. A Hierarchical shape descriptions was widely used in sketch systems \cite{HierarchicalParsing7,SketchRead2007}.  %Ladder30,AlvaradoFreedom42

Recognition based on AI methods as applied in \cite{SketchRead2007} has the principle problem of computational complexity of the matching between shape description and geometric primitives produced in the segmentation process. Natural sketch are difficult to be segmented and grouped to create reliable fragment of geometric primitives. This is partially due to the increase of noise and other phenomena such as overlapped, over traced strokes and sharing of strokes between different symbols. Since these systems are dependable on the segmentation results, they either perform recognition using undependable segmentation results or consider different ways to segment the input sketch. This is because they assume that each stroke plays a role in the structure of the shape. 
%The role of which each stroke plays in the structure of the sketched symbol is used to determine shapes in stroke based methods.  
%Gestures are pen strokes drawn by the user that can immediately recognized by the system.
%role of which each stroke plays in the structure of the sketched symbol is used to determine shapes 
%. Describing the role  how user draws individual strokes  properties that describe how users draw individual strokes in drawing the symbol.  to. The identification of individual stroke is used to determine The symbols are recognized by identifying the individual strokes.
In PR based methods, systems compute properties that describe symbols then use a classification method to recognize symbols. The properties used can either be based on gesture properties, global shape properties or image based properties. Gesture properties are set of features that represent how the user draws a symbol not how it looks like (e.g., identify a rectangle when a user draws a L corner). In gesture properties approach, systems identify symbols by using gesture properties of some of its underlying strokes. System that uses gestures properties assume that users draw symbols in specific order and style \cite{gestureexample12,aideddesgin22}. In global shape properties approach, systems investigate general shape properties and its underlying strokes. Using shape properties (e.g., ratio of the area of convex hull to the area of shape) relaxes the assumption and the role of each stroke in the symbol but it does not truly represent the appearance or fine details of the symbol \cite{DiagramOfflineConvexHull,Cali63}. Unlike shape properties based methods, image based methods disregard individual strokes and focus on the overall appearance of shapes. These system convert the shapes into image and extracted image based features (e.g. local gradient feature)\cite{Oltmans07,imagetrainable48}. In \cite{imagetrainable48}, they used four different distance measures to match user symbol with the prototype database. The prototype of a shape is represented as an image of the symbol that is normalized and scaled.  However, the problem is that the shape prototype must represent all transformation and variation the input shape can have.   %Image based methods are used in \cite{Oltmans07,imagetrainable48} where they computed a code book of 1000 features to recognize symbols.

@@

In this paper, we choose to implement a hybrid method between shape property, gesture and hierarchical (AI method) based approaches. Our method attempts to avoid many of the problems faced by approaches based on individual gestures or global shape properties while trying to gain the additional information that the stroke based system uses. Firstly, the system uses particle swarm optimization (PSO) to optimally segment user's strokes. Users can draw symbols using any number of strokes in any order they like. The system segments each stroke using two different PSO algorithms. The first segmentation algorithm is based on dividing the digital curve into polygons\cite{PolygonApproximationPSO}. Our algorithm enhances its performance by adding curvature and other local information while segmenting strokes to achieve better results. To handle complex strokes, our system uses a second segmentation PSO algorithm to segment strokes into a set of lines and circular arcs. After segmentation, a feature vector composed of different statistical, geometrical and composite features is deduced. The final classification step uses a SVM classifier to classify symbols into one of the previously trained classes based on the features vector computed.

 %In general, the appearance based methods use image processing techniques to recognize sketches that the users draw. 
%Sketch recognition can be based on: strokes, global shape properties, and appearance \cite{Oltmans07}. The role of which each stroke plays in the structure of the sketched symbol is used to determine shapes in stroke based methods. Most stroke based systems either represent symbols as gestures or as a hierarchical shape. Gestures are pen strokes drawn by the user that can immediately recognized by the system. A Hierarchical shape descriptions was widely used in sketch systems \cite{HierarchicalParsing7,Ladder30,AlvaradoFreedom42,SketchRead2007}. Symbols are represented using hierarchical levels of simpler geometric shapes or primitives and spatial relations between them. Systems that employ such representation must first pre-process strokes to break them into geometric primitives. Recognition can then be treated as sub graph matching problem.

@@

\subsection{Preprocessing}
\label{Prepross}% Agar \cite{polygonfeedback31} mentions that the pointing device (for example the mouse) sampling rate is the reason for this phenomena.
 Speed and time difference information were widely used in sketch recognition systems \cite{earlyprocess}. Figure \ref{fig:orignalStroke} shows an example of an input stroke drawn by users.  Figure \ref{fig:speed2Distance} shows the graph of time difference and the speed for the stroke in Figure \ref{fig:orignalStroke}. It was noted that time difference provides more distinct maxima than the speed information. Agar et al. \cite{polygonfeedback31} mentions that the pointing device (e.g., mouse) sampling rate is the reason for this phenomenon. The points are sampled at regular time intervals as long as the pointing device is moving. Therefore, there are no new detected points when the pointing device is stationary. This leads to a nearly constant time difference between samples while the pen is moving and large difference while the pen is stationary (Figure \ref{fig:timediff}). Contrary to time difference information, the speed information, where the users draw with variant speeds, the speed information has a lot of noise in the data (Figure \ref{fig:speed2Distance}).  %Similarly, direction information is used as it provides better distinctive maxima for the corners than the curvature information \cite{meanshift10}.% Figure \ref{fig:curvatures} shows the direction and curvature graphs for the stroke drawn in Figure \ref{fig:orignalStroke}

@@

 \begin{figure}
	\centering
			\subfigure[Speed versus Displacement Graph]{	\includegraphics[scale=0.4]{images/vel3.jpg}}
			\hfill
			\subfigure[ Time Difference  versus Displacement Graph] {\label{fig:timediff}\includegraphics[scale=0.4]{images/td3.jpg}}
	\caption{Speed and Time Difference Graphs}  a) speed graph has lot of noise in the data with respect to b) time difference.   The black points denote the location of the possible dominant points $P_{pd}$. Black rectangles represent regions $R_i$ that in a) Are lower than the threshold. b) Are above the threshold. 
	\label{fig:speed2Distance}
\end{figure}
%is calculated as the angle between two vectors and  where $Q_i=\overrightarrow {P_{i - 1} P_i }$ is the vector from the point $P_{i - 1}$ to point $P_i$ 
Dominant points are characterized by low speed values, high curvature and direction values. However, the dominant points from the curvature information may not be the same points as the dominant points extracted from the speed information. Hence, we compute time difference, direction, speed and curvature of each point along a stroke to ensure that no dominant point is overlooked in this phase. The speed is calculated as $v=\Delta s/\Delta t$ where $t$ is the time difference between two points and $s$ is the length between them. The direction $d$ calculated as the angle between vectors $\overrightarrow {P_{i - 1} P_i }$ and the $x-axis$ and the curvature is considered as the change in direction $d$ with respect to length $s$ i.e. $c= \Delta d/\Delta s$.

@@

\subsection{Segmentation}
\label{seg}
In the segmentation stage, the input stroke is divided into a set of primitives. As shown in Figure \ref{fig:segblock} first an attempt is made to fit the stroke points into a curve or an ellipse using a minimum square error fitting algorithm \cite{ellipsefit}. The ellipse fitting step mainly helps to prevent the system from over segmenting the stroke into multiple lines or curves if the input stroke is an ellipse. If the stroke proved to be an ellipse arc then the segmentation process ends and the system proceeds to the next step. Otherwise, the stroke is passed to two further segmentation algorithms that divide the stroke to either lines or lines and curves. Hence we generate two different segmentations; the system chooses the segmentation with the minimum error. The following sections describe in details the ellipse detection algorithm and the two segmentation algorithms used to divide the stroke. % The input to these algorithms are the stroke points along with the possible dominant points $P_{pd}$ computed during preprocessing. The output is a set of dominant points which are connected by either lines or curves. 
%After two different segmentations are generated, 
%///AllBlockDiagram.jpg  you can change the block to the main block. 
 \begin{figure}
	\centering
		\includegraphics[scale=0.48]{images/blockSmall.jpg}
	\caption{Segmentation Block} 
	\label{fig:segblock}
\end{figure}
\subsection{Ellipse fitting}
%The process starts by computing the center of the stroke bounding box. The bounding box center point is used as the first estimation of the ellipse center. The axes of the ellipse are estimated as $width/2$ and $height/2$ of the stroke bounding box. The least square fitting algorithm is used to minimize the fitting error of the ellipse equation \cite{chernov}.  
This process attempts to fit the stroke points into an ellipse arc; it starts with computing the center of the stroke bounding box. The bounding box center point is used as the first estimation of the center of the ellipse. The axes of the ellipse are estimated as the $width/2$ and $height/2$ of the stroke bounding box where $width$ and $height$ are the width and height of the bonding box. The least square fitting algorithm \cite{chernov} is used to minimize the fitting error of the ellipse Equation (\ref{eq:circleFit})  
\begin{equation}
E = \sum\limits_{i = 0}^N {\frac{{(x_i - x_0 )}}{{a^2 }}^2  + \frac{{(y_i - y_0 )}}{{b^2 }}^2  - 1} 
\label{eq:circleFit}
\end{equation}
where $N$ is number of points in the stroke, $a$ \& $b$ are the length of ellipse axes, $x_0$ \& $y_0$ are the coordinates of the center point and $x_i$ \& $y_i$ are the coordinates of point $i$ in the stroke. A list of new values for $x_0$ , $y_0$ ,$a$ and $b$ are generated randomly from the older values with small increments after each loop.  After few iterations, the final fit error of the estimated ellipse is reported. Another measure is used to compute the efficiency of the final estimated ellipse. Equation \ref{eq:circleError} ensures that the drawn percentage of ellipse is considered. This eliminates fitting a line into very large ellipse but leaves small ellipses to be fitted as a partial or a full ellipse.

@@

After each iteration of the swarm algorithm (\textsl{AlgS1} and \textsl{AlgS2}), each particle is refined using the following procedures: 
\begin{enumerate}
	\item For each particle $P_i$ each dominant point $P_{ij}$ is checked to find if it was labeled before as a \textit{possible dominant point} $P_{pd}$ (computed as in section \ref{Prepross}). If it was not labeled, the point $P_{ij}$ is moved to the nearest labeled point. This ensures that all of the points generated by the DPSO are possible dominant points $P_{pd}$. 
	\item The particles are tested to make sure that the distance between every two successive dominate point is larger than $min_D$, where $min_D$ is 5\% of the total length of the stroke.  Otherwise, two temperary particles are evaluated by removing either point from the current particle. After that, one of these points is removed based on the error caused by its removal.
	\item Each two successive segments are tested to determine if they can be merged into a single segment. If merging the two segments results in segmentation that has an error value less than or equal to the error before merging, the two segments are merged. This step ensures that the final segmentation of the system has minimum number of segments. 
	% the same or smaller the error
\end{enumerate}
%1) 2)  % If two points are nearer than $min_D$ then one of the points is removed. 
\subsection{Recognition}
\label{sec:Recognition}
After the user draws all strokes of the symbol, the set of un-recognized strokes is grouped together along with their segmentation as input to the feature extraction process. A composite set of features is used to generate a single feature vector. The following list describes each feature in details.
%\begin{description}
\subsubsection{Structural and geometrical Features (FS1)}
  They are features defines the structure of the geometrical symbol such as;   
%\begin{description}
%	 \item[No. Of segments] Number of segments in the symbol. 
%		\item[No. Of primitives] Number of primitives in the symbol. The feature helps when identifying             symbols with mixed geometric primitives like cylinders and callouts.  
%		\item [No. of curves] Number of curves or ellipses in the symbol. 
%		\item [No. of lines] Number of lines in the symbol. 
%		\item [No. of perpendicular lines] Number of perpendicular lines. 
%		\item [No. of parallel lines] Number of parallel lines. 
%		\item [No. of intersections] Number of intersection between lines and curves. 
%		\item [Size Ratio] Ratio between width to height of the symbol.
%		%\item[No. of holes] Number of holes in the symbols. 
%
%\end{description}
 \begin{itemize}
	 \item \emph{Segments:} The number of segments in the symbol.
	 \item \emph{Strokes:} The number of strokes or partial strokes that created the symbols.  
		\item  \emph{Primitives:} The number of primitives in the symbol. This feature helps when identifying  symbols with mixed geometric primitives such as cylinders and callouts.  
		\item \emph{Curves:} The number of curves or ellipses in the symbol. 
		\item \emph{Lines:} The number of lines in the symbol. 
		\item \emph{Perpendicular}\emph{lines:} The number of perpendicular lines. 
		\item \emph{Parallel}\emph{lines:} The number of parallel lines. 
		\item \emph{Intersections:} The number of intersection between lines and curves. 
		\item \emph{T intersections:} The number of T intersections. 
		\item \emph{L intersections:} The number of L intersections. 
		\item \emph{X intersections:} The number of X intersections.	
		%\item[No. of holes] Number of holes in the symbols. 
\end{itemize}
\subsubsection{Rubine Features (FS2)}
  These features introduced by Rubine \cite{gestureexample12} for single stroke gestures. Such as length of stroke, cosine of starting angle, smoothness of symbol and the properties of the bounding box of the symbol.
